---
title: "🗺️ Interactive Mapping with mapgl and Shiny"
author: "DS421 – Cartographic Design"
format:
  html:
    code-fold: true
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

## 1️⃣ Census Shiny App

#### Load Libraries

```{r}
library(shiny)
library(tidyverse)
library(sf)
library(mapgl)
library(lubridate)
library(bslib)
library(here)
library(tidycensus)
library(ggpubr)
library(plotly)
library(mapview)
library(leaflet)

# Read the fire data
fire_data <- read_csv("LANDSAT_USA_contiguous_and_Hawaii_7d.csv")

# Check if it loaded
head(fire_data)
```

#### View Census Variables

```{r}
usa <- map_data("state")
```

#### Transform Data

```{r}
# Define raw ACS variable codes
raw_vars <- c(
  median_income = "B19013_001",
  bachelors_count = "B15003_022",  # Count, not percent
  median_gross_rent = "B25064_001",
  total_population = "B01003_001"
)

# Labels for Shiny dropdown (key = column name in wide data)
variable_choices <- c(
  "Median Household Income" = "median_income",
  "Percent with Bachelor's Degree" = "pct_bachelors",
  "Median Gross Rent" = "median_gross_rent"
)

# Pull and transform ACS data
hawaii_data_raw <- get_acs(
  geography = "tract",
  state = "HI",
  variables = raw_vars,
  geometry = TRUE,
  year = 2023
) |>
  filter(GEOID != "15003981200") |>  # Remove Kalawao County
  st_transform(4326)

# Convert to wide format + calculate % with bachelor's
hawaii_data <- hawaii_data_raw |>
  select(GEOID, NAME, variable, estimate, geometry) |>
  st_drop_geometry() |>   # temp drop for pivot
  pivot_wider(names_from = variable, values_from = estimate) |>
  mutate(
    pct_bachelors = 100 * bachelors_count / total_population
  ) |>
  left_join(
    hawaii_data_raw |> select(GEOID, geometry) |> distinct(),
    by = "GEOID"
  ) |>
  st_as_sf()
```

## 🧩 Challenge 1

Can you make a histogram of median rent colored by county. Hint: You will have to make a new column called county based on the values in the NAME column

```{r}
# Create county column from NAME
hawaii_data <- hawaii_data |>
  mutate(
    county = case_when(
      str_detect(NAME, "Hawaii County") ~ "Hawaii",
      str_detect(NAME, "Honolulu County") ~ "Honolulu",
      str_detect(NAME, "Kauai County") ~ "Kauai",
      str_detect(NAME, "Maui County") ~ "Maui",
      TRUE ~ "Other"
    )
  )

```

```{r}
# Create the histogram
ggplot(hawaii_data, aes(x = median_gross_rent, fill = county)) +
  geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of Median Gross Rent by County",
    x = "Median Gross Rent ($)",
    y = "Count of Census Tracts",
    fill = "County",
    caption = "Source: ACS 2023 5-year estimates"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )
```

## 🧩 Challenge 2

A. Can you make a scatterplot showing the relationship between median income and median gross rent. Can you use the `stat_cor()`function to show a correlation coefficient and p value

```{r}
ggplot(hawaii_data, aes(x = median_income, y = median_gross_rent)) +
  geom_point(alpha = 0.6, color = "darkblue", size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  stat_cor(
    method = "pearson",
    label.x = 50000,
    label.y = 2500,
    size = 5
  ) +
  labs(
    title = "Relationship between Median Income and Median Gross Rent",
    x = "Median Household Income ($)",
    y = "Median Gross Rent ($)",
    caption = "Source: ACS 2023 5-year estimates"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

```{r}

```

B. Can you make this scatterplot interactive with `ggplotly()`

```{r}


p <- ggplot(hawaii_data, aes(
  x = median_income,
  y = median_gross_rent,
  text = paste(
    "County:", county, "<br>",
    "Income: $", format(median_income, big.mark = ","), "<br>",
    "Rent: $", format(median_gross_rent, big.mark = ",")
  )
)) +
  geom_point(alpha = 0.6, color = "darkblue", size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  stat_cor(method = "pearson") +
  labs(
    title = "Relationship between Median Income and Median Gross Rent",
    x = "Median Household Income ($)",
    y = "Median Gross Rent ($)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

ggplotly(p, tooltip = "text")
```

## 🧩 Challenge 3

Can you make a choropleth map of Hawaii using `mapview()` that shows median rent by census tract

```{r}
mapview(
  hawaii_data,
  zcol = "median_gross_rent",
  layer.name = "Median Gross Rent",
  col.regions = viridisLite::viridis(10),
  alpha.regions = 0.7,
  legend = TRUE,
  map.types = "CartoDB.Positron"
)
```

#### Building a Shiny App

We have these 3 variables in hawaii_data, it would be nice if users can select a variable from a dropdown: [Introducing Shiny](https://shiny.posit.co/)

There are many mapping packages we can use with Shiny. The simplest for now is `mapview`. It is good to note that the mapgl package is also designed specifically to [work with advanced shiny apps](https://walker-data.com/mapgl/articles/shiny.html).

```{r}

# Define choices
variable_choices <- c(
  "Total Population" = "total_pop",
  "Median Income" = "median_income",
  "Housing Units" = "housing_units",
  "Land Area" = "land_area"
)

ui <- fluidPage(
  titlePanel("Hawaii Census Map"),
  sidebarLayout(
    sidebarPanel(
      selectInput("var", "Choose a variable to map:", choices = variable_choices)
    ),
    mainPanel(
      leafletOutput("map", height = "700px")
    )
  )
)

server <- function(input, output, session) {
  output$map <- renderLeaflet({
    # Get the selected variable
    selected_var <- hawaii_data[[input$var]]
    
    # Create color palette
    pal <- colorNumeric("viridis", selected_var)
    
    leaflet(hawaii_data) |>
      addProviderTiles("CartoDB.Positron") |>
      addPolygons(
        fillColor = ~pal(selected_var),
        weight = 1,
        opacity = 1,
        color = "white",
        fillOpacity = 0.7,
        label = ~paste(names(variable_choices)[which(variable_choices == input$var)], ":", 
                       format(selected_var, big.mark = ",")),
        labelOptions = labelOptions(direction = "auto")
      ) |>
      addLegend(
        pal = pal,
        values = selected_var,
        title = names(variable_choices)[which(variable_choices == input$var)],
        position = "bottomright"
      )
  })
}

shinyApp(ui, server)
```

## 🧩 Challenge 4

Can you create a shiny app using your LANDSAT_USA_contiguous_and_Hawaii_7d.csv in the data folder that allows users to filter the map by confidence. For example, the user can select from a side dropdown to only show *high* confidence fires, *medium* confidence fires, or *low* confidence fires. Do not use AI for this and use the example above. AI Bonus: Add a date slider so users can toggle the map by date. Can you add a heatmap layer users can toggle on and off?

```{r}

ui_fire <- fluidPage(
  titlePanel("🔥 Wildfire Confidence Explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput(
        "confidence_filter",
        "Filter by Confidence Level:",
        choices = c("All", "high", "medium", "low"),
        selected = "All"
      ),
      
      # Bonus: Date range slider
      sliderInput(
        "date_range",
        "Select Date Range:",
        min = as.Date(min(fire_data$acq_date)),
        max = as.Date(max(fire_data$acq_date)),
        value = c(as.Date(min(fire_data$acq_date)), as.Date(max(fire_data$acq_date))),
        timeFormat = "%Y-%m-%d"
      ),
      
      # Bonus: Heatmap toggle
      checkboxInput(
        "show_heatmap",
        "Show Heatmap",
        value = FALSE
      ),
      
      hr(),
      p("Data source: NASA FIRMS LANDSAT")
    ),
    mainPanel(
      leafletOutput("fire_map", height = "700px")
    )
  )
)

server_fire <- function(input, output, session) {
  
  filtered_data <- reactive({
    df <- fire_sf
    
    if (input$confidence_filter != "All") {
      df <- df |> filter(confidence == input$confidence_filter)
    }
    
    df <- df |>
      filter(
        as.Date(acq_date) >= input$date_range[1],
        as.Date(acq_date) <= input$date_range[2]
      )
    
    df
  })
  
  output$fire_map <- renderLeaflet({
    df <- filtered_data()
    
    if (nrow(df) == 0) {
      return(
        leaflet() |>
          addProviderTiles("CartoDB.DarkMatter") |>
          setView(lng = -98, lat = 39, zoom = 4) |>
          addControl("No fires match the selected criteria", position = "topcenter")
      )
    }
     map <- leaflet(df) |>
      addProviderTiles("CartoDB.DarkMatter") |>
      setView(lng = -98, lat = 39, zoom = 4)
    
    if (input$show_heatmap) {
      map |> 
        addHeatmap(
          lng = ~st_coordinates(geometry)[,1],
          lat = ~st_coordinates(geometry)[,2],
          blur = 20,
          max = 0.05,
          radius = 15
        )
    } else {
      map |>
        addCircleMarkers(
          color = ~case_when(
            confidence == "high" ~ "#FF0000",
            confidence == "medium" ~ "#FFA500",
            confidence == "low" ~ "#008000"
          ),
          radius = 3,
          fillOpacity = 0.7,
          opacity = 0.8,
          popup = ~paste(
            "<b>Fire Detection</b><br>",
            "Date:", acq_date, "<br>",
            "Time:", acq_time, "<br>",
            "Satellite:", satellite, "<br>",
            "Confidence:", confidence
          ),
          label = ~paste("Fire -", confidence),
          labelOptions = labelOptions(
            style = list("font-weight" = "normal", padding = "3px 8px"),
            textsize = "12px",
            direction = "auto"
          )
        ) |>
        addControl(
          html = "<div style='background-color:white;padding:8px 12px;border-radius:5px;box-shadow:0 0 10px rgba(0,0,0,0.3);'>
                  <b>🔥 Fire Confidence</b><br>
                  <span style='color:#FF0000;'>●</span> High<br>
                  <span style='color:#FFA500;'>●</span> Medium<br>
                  <span style='color:#008000;'>●</span> Low
                  </div>",
          position = "bottomright"
        )
    }
  })
}

# Uncomment to run the app
# shinyApp(ui_fire, server_fire)
   
```

```{r}

```

## 🧩 Challenge 5

Can you add the functionality to your census shiny app (new panel) that allows users to pick two census variables to plot on an interactive scatter plot using `ggplotly()` and `stat_cor()`

We should remain careful about our use/potential dependence on AI for coding but for cases like these where we have built all the individual parts, AI can be really helpful in putting them together in a sleek design on a shiny app.

```{r}
# Challenge 5: Enhanced Census App with Scatter Plot

ui_census_enhanced <- fluidPage(
  titlePanel("🌺 Hawaii Census Explorer"),
  
  tabsetPanel(
    # Map Tab
    tabPanel(
      "🗺️ Map",
      sidebarLayout(
        sidebarPanel(
          selectInput(
            "var_map",
            "Choose a variable to map:",
            choices = variable_choices
          )
        ),
        mainPanel(
          leafletOutput("census_map", height = "700px")
        )
      )
    ),
    
    # Scatter Plot Tab
    tabPanel(
      "📊 Scatter Plot",
      sidebarLayout(
        sidebarPanel(
          selectInput(
            "var_x",
            "Choose X-axis variable:",
            choices = c(
              "Median Income" = "median_income",
              "Median Gross Rent" = "median_gross_rent",
              "Percent Bachelor's" = "pct_bachelors"
            ),
            selected = "median_income"
          ),
          selectInput(
            "var_y",
            "Choose Y-axis variable:",
            choices = c(
              "Median Gross Rent" = "median_gross_rent",
              "Median Income" = "median_income",
              "Percent Bachelor's" = "pct_bachelors"
            ),
            selected = "median_gross_rent"
          )
        ),
        mainPanel(
          plotlyOutput("scatter_plot", height = "600px")
        )
      )
    )
  )
)

server_census_enhanced <- function(input, output, session) {
  
  output$census_map <- renderLeaflet({
    mapview(
      hawaii_data,
      zcol = input$var_map,
      layer.name = names(variable_choices)[variable_choices == input$var_map],
      col.regions = viridisLite::viridis(10),
      alpha.regions = 0.7
    )@map
  })
  
  output$scatter_plot <- renderPlotly({
    df <- hawaii_data |> st_drop_geometry()
    
    p <- ggplot(df, aes(x = .data[[input$var_x]], 
                        y = .data[[input$var_y]],
                        text = paste(
                          "County:", county, "<br>",
                          names(variable_choices)[variable_choices == input$var_x], ": $", 
                          format(.data[[input$var_x]], big.mark = ","), "<br>",
                          names(variable_choices)[variable_choices == input$var_y], ": $", 
                          format(.data[[input$var_y]], big.mark = ",")
                        ))) +
      geom_point(alpha = 0.6, color = "darkblue", size = 2) +
      geom_smooth(method = "lm", se = TRUE, color = "red", alpha = 0.2) +
      stat_cor(
        method = "pearson",
        label.x = min(df[[input$var_x]], na.rm = TRUE),
        label.y = max(df[[input$var_y]], na.rm = TRUE),
        size = 4
      ) +
      labs(
        title = paste(
          "Relationship between",
          names(variable_choices)[variable_choices == input$var_x],
          "and",
          names(variable_choices)[variable_choices == input$var_y]
        ),
        x = names(variable_choices)[variable_choices == input$var_x],
        y = names(variable_choices)[variable_choices == input$var_y]
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold")
      )
    
    ggplotly(p, tooltip = "text")
  })
}

# Uncomment to run the app
# shinyApp(ui_census_enhanced, server_census_enhanced)
```
